AFQ-Insight

### 2019-12-12 reports/manuscript/sections/conclusion.tex

Neuroscience has entered an era in which consortium efforts are putting together
large datasets of high-quality dMRI measurments to address a variety of
scientific questions \cite{jernigan2016ping, jernigan2018abcd,
alexander2017open, Miller2016-hw, VanEssen2012}. The volume and complexity of
these data pose a substantial challenge. Dimensionality reduction with
tractometry, followed by analysis with the approach we present here promises to
capitalize on the wealth of data and on the co-measurement of interesting and
important phenotypical data about brain health and about the participants'
cognitive abilities. We also expect the group-regularized approach to improve
with larger datasets.

>>>>

Neuroscience has entered an era in which consortium efforts are putting together
large datasets of high-quality dMRI measurements to address a variety of
scientific questions \cite{jernigan2016ping, jernigan2018abcd,
alexander2017open, Miller2016-hw, VanEssen2012}. The volume and complexity of
these data pose a substantial challenge. Dimensionality reduction with
tractometry, followed by analysis with the approach we present here promises to
capitalize on the wealth of data and on the co-measurement of interesting and
important phenotypical data about brain health and about the participants'
cognitive abilities. We also expect the group-regularized approach to improve
with larger datasets.

### 2019-12-12 reports/manuscript/sections/introduction.tex

Diffusion-weighted Magnetic Resonance Imaging (dMRI) provides a unique view into
the physical properties of the connections that comprise the brain white matter.
While the measurements are usually conducted with voxels at the millimeter
scale, water molecules within each voxel probe the structure of white matter at
the micrometer scale providing important information about the physical
structure of the white matter, including  the distribution of fiber orientations
and density of axons within each voxel \cite{wandell2016clarifying}. Even though
metrics derived from diffusion measurements are ambiguous in terms of their
underlying biological interpretation, analyzing the variance in these properties
has proven useful in characterizing individual differences in cognitive
function, characterizing differences between populations and detecting
abnormalities associated with disease \cite{Thomason2011-qn}

To relate the diffusion in each voxel to the macro-structure of long-range
connections between different brain regions, methods for computational
tract-tracing from diffusion MRI, or tractography, combine the estimates of
fiber orientations in each voxel to form streamlines that traverse the volume of
the white matter \cite{Conturo1999-je, Mori2002-qi}. These methods have been
under increased scrutiny and several lines of investigation have raised
critiques of their validity \cite{Maier-Hein2017-vb, Thomas2014-ki}. On the
other hand, there have been efforts to shore up the inferences made with these
methods \cite{Pestilli2014NatMeth, Takemura2016-sh, Smith2013-nc, Smith2015-cx,
Smith2015-zt, Rheault2018-wk}. Importantly, though discovering novel fiber
bundles requires extraordinary evidence, and delineating the exact cortical
termination of the streamlines in the gray matter is still prone to error, there
is little dispute that tractography can accurately define the location of major
white matter bundles that are known to exist within the core of the white matter
\cite{Maier-Hein2017-vb}.

Using this fact, one of the most powerful methods currently available to put
macro- and micro-structure together is \emph{tractometry} (also known as
tract-based morphometry): assessment of the physical properties of the white
matter along specific tracts \cite{Bells2011-cf}. Though there are several
different available implementations of this overall idea, the principles are
similar \cite{yeatman2012tract, Yendiki2011-ay, Wassermann2016-iv, ODonnell2009-uu}:
tractometry begins by delineating the parts of the white matter that belong to
different major tracts or axonal fiber bundles, such as the corticospinal tract
or arcuate fasciculus, and samples the biophysical properties (such as
fractional anisotropy or mean diffusivity) along the length of these bundles. In
some cases, tissue properties along the length of each bundle are summarized by
taking the mean along each bundle, but there is a large body of evidence showing
that there is systematic variability within each bundle. This justifies
retaining the individual samples along the length of each fiber
\cite{yeatman2012tract, colby2012, ODonnell2009-uu}. While this retains
important information about each individual's white matter, it also presents
statistical challenges due to the dimensionality of the data. Based on
tractometry, researchers may choose to compare different individuals to each
other. This is usually done following one of the following approaches:

@@

\item Mass univariate approaches: In this approach comparisons between groups or
across individuals are done at each node of each tract, for each one of the
diffusion metrics available at that point. This approach is exhaustive, but the
challenge with this approach is that the are multiple points to compare, and
statistical power is compromised by a multiple comparison problem. Different
approaches can be taken to resolving this challenge. For example, Colby and
colleagues \cite{colby2012} used a non-parametric resmapling approach to correct
for family-wise error across the universe of different comparisons available in
the mass univariate approach \cite{Nichols2002-zu, Nichols2003-yy}.

\item ROI-based approach: An alternative is using an ROI based approach. In this
approach, the multiple comparison problem is circumvented by selecting just a
few tracts to compare in each individual, or even focusing on particular
segments of these tracts based on \emph{a priori} hypotheses. This approach is
very powerful when the biological basis of the process of interest is relatively
well understood (for a recent example, see \cite{huber2018rapid}).

\item ROI-based selection, followed by multivariate analysis. Here, an ROI is
selected based on \emph{a priori} knowledge, and all the nodes or voxels in the
ROI are used together to fit a model that can predict differences between
individuals. An example of that is the ``profilometry'' framework, in which
different diffusion metrics from a single tract are combined together to provide
input to a multivariate analysis of covariance, and linear discriminant analysis
\cite{dayan2016profilometry}.

@@

Ideally, we would like a method that simultaneously capitalizes on all the data
across all tracts to make the best possible prediction, while also retaining and
surfacing spatial information about the locations that are most informative for
a prediction. In the present work, we developed a novel framework for analysis
of tractometry that simultaneously selects the features for analysis, and fits a
model to these features. Rather than a comparative approach, this framework
takes a predictive approach to inference \cite{Breiman2001-uz}. We use a linear
modeling approach, which aims to predict phenotypical variance in a group of
subjects, based on a linear combination of the features estimated with
tractometry. Using this approach, we need to first deal with the large and
asymmetric dimensionality of the data: tractometry data usually has many more
features (i.e., number of measurements per individual) than samples (number of
subjects), which makes inferences from the data about phenotypical differences
between individuals ill-posed. This regime is the target of several statistical
learning techniques, and is often solved by various forms of regularization. For
example, Tikhonov regularization shrinks the solution such that the sum of
squared contributions from the individual features are minimized
\cite{Hoerl2000-ij}. Another solution to the problem is provided by the Lasso
algorithm, which instead minimizes the sum of the absolute values of
contributions of each feature \cite{Tibshirani1996-qs}. This tends to shrink to
zero the contributions of many of the features, providing results that are both
accurate and interpretable. When additional structure is available in the
organization of the data, regularization algorithms can take advantage of this
structure. For example, if the features lend themselves to a natural division
into different groups, the group lasso (GL) can be used to select groups of
features, rather than individual features \cite{Yuan2006-ky}. The Sparse Group
Lasso (SGL) elaborates on this idea by providing control both of group sparsity,
as well as overall sparsity of the solutions \cite{simon2013sgl}. Because the
features measured with tractomery lend themselves to grouping based on the
tracts from which each measurement is taken, GL and SGL provide a useful tool
linear model fitting in problems of this form. Using linear regression as the
basis for analysis is useful for both regression tasks, in which a continuous
measure (e.g., age) is modeled, as well as for classification tasks, where a
discrete property is modeled (e.g., whether an individual has a particular
disease). Moreover, use of SGL helps derive informative properties of the human
white matter, both in cases where the underlying causes of variance are sparsely
concentrated (e.g., in one particular white matter tract) and in cases where
biological substrates of differences between individuals are more widespread.
This is because SGL is controlled by meta-parameters that set the degree to
which group sparsity and overall sparsity are applied. These metaparameters can
be set empirically, based on the data, through a process of nested
cross-validation. We demonstrate these properties of the algorithm in data from
previously published studies \cite{sarica2017corticospinal,
yeatman2014lifespan}.

>>>>

Diffusion-weighted Magnetic Resonance Imaging (dMRI) provides a unique view into
the physical properties of the connections that comprise the brain white matter.
While the measurements are usually conducted with voxels at the millimeter
scale, water molecules within each voxel probe the structure of white matter at
the micrometer scale providing important information about the physical
structure of the white matter, including the density of axons and distribution
of nerve fascicle orientations within each voxel \cite{wandell2016clarifying}.
Even though metrics derived from diffusion measurements are ambiguous in terms
of their underlying biological interpretation \cite{Jones2013-xv}, analyzing the
variance in these properties has proven useful in characterizing individual
differences in cognitive function, characterizing differences between
populations and detecting brain abnormalities associated with disease
\cite{Thomason2011-qn}.

To relate the diffusion in each voxel to the macro-structure of long-range
connections between different brain regions, methods for computational
tract-tracing from diffusion MRI, or tractography, combine the estimates of
fascicle orientations in each voxel to form streamlines that traverse the volume
of the white matter \cite{Conturo1999-je, Mori2002-qi}. These methods have been
under increased scrutiny and several lines of investigation have raised
critiques of their validity \cite{Maier-Hein2017-vb, Thomas2014-ki}. On the
other hand, there have been efforts to shore up the inferences made with these
methods \cite{Pestilli2014NatMeth, Takemura2016-sh, Smith2013-nc, Smith2015-cx,
Smith2015-zt, Rheault2018-wk}. Importantly, though discovering novel nerve fiber
bundles requires extraordinary evidence, and delineating the exact cortical
termination of the streamlines in the gray matter is still prone to error, there
is little dispute that tractography can accurately define the location of
several major white matter bundles that are known to exist within the core of
the white matter \cite{Maier-Hein2017-vb}.

Using this fact, one of the most powerful methods currently available to put
macro- and micro-structure together is \emph{tractometry} (also known as
tract-based morphometry): assessment of the physical properties of the white
matter along specific tracts \cite{Bells2011-cf}. Though there are several
different available implementations of this overall idea, the principles are
similar \cite{yeatman2012tract, Yendiki2011-ay, Wassermann2016-iv,
ODonnell2009-uu}: tractometry begins by delineating the parts of the white
matter that belong to different major tracts or axonal nerve fiber bundles, such
as the corticospinal tract or arcuate fasciculus, and samples the biophysical
properties (such as fractional anisotropy or mean diffusivity) along the length
of these bundles. In some cases, tissue properties along the length of each
bundle are summarized by taking the mean along each bundle, but there is a large
body of evidence showing that there is systematic variability within each
bundle. This justifies retaining the individual samples along the length of each
bundle \cite{yeatman2012tract, colby2012, ODonnell2009-uu}. While this retains
important information about each individual's white matter, it also presents
statistical challenges due to the dimensionality of the data. Based on
tractometry, researchers may choose to compare different individuals to each
other. This is usually done following one of the following approaches:

@@

\item Mass univariate approaches: In this approach comparisons between groups or
across individuals are done at each node of each tract, for each one of the
diffusion metrics available at that point. This approach is exhaustive, but
statistical power is compromised by a multiple comparison problem. Different
approaches can be taken to resolving this challenge. For example, Colby and
colleagues \cite{colby2012} used a non-parametric resmapling approach to correct
for family-wise error across the different possible comparisons
\cite{Nichols2002-zu, Nichols2003-yy}.

\item Region of interest(ROI)-based approaches: An alternative that circumvents
the multiple comparison problem is to select just a few tracts to compare in
each individual, or even focusing on particular segments of these tracts based
on \emph{a priori} hypotheses. This approach is very powerful when the
biological basis of the process of interest is relatively well understood (for a
recent example, see \cite{huber2018rapid}).

\item ROI-based selection, followed by multivariate analysis: Here, an ROI is
selected based on \emph{a priori} knowledge, and all the nodes or voxels in the
ROI are used together to fit a model that can predict differences between
individuals. An example of that is the ``profilometry'' framework, in which
different diffusion metrics from a single tract are combined together to provide
input to a multivariate analysis of covariance, and linear discriminant analysis
\cite{dayan2016profilometry}.

@@

Generally speaking analysis methods should balance predictive accuracy with
descriptive power \cite{Murdoch2019-ax, Breiman2001-uz}. Accordingly,
tractometry analysis should simultaneously capitalize on all the data across all
tracts to make the best possible prediction, while also retaining and surfacing
spatial information about the locations that are most informative for a
prediction. In the present work, we developed a novel framework for analysis of
tractometry that simultaneously selects the features for analysis, and fits a
model to these features. We use a linear modeling approach, which aims to
predict phenotypical variance in a group of subjects, based on a linear
combination of the features estimated with tractometry. Using this approach, we
need to first deal with the large and asymmetric dimensionality of the data:
tractometry data usually has many more features (i.e., number of measurements
per individual) than samples (number of subjects), which makes inferences from
the data about phenotypical differences between individuals ill-posed. This
regime is the target of several statistical learning techniques, and is often
solved by various forms of regularization. For example, Tikhonov regularization
shrinks the solution such that the sum of squared contributions from the
individual features are minimized \cite{Hoerl2000-ij}. Another solution to the
problem is provided by the Lasso algorithm, which instead minimizes the sum of
the absolute values of contributions of each feature \cite{Tibshirani1996-qs}.
This tends to shrink to zero the contributions of many of the features,
providing results that are both accurate and interpretable. When additional
structure is available in the organization of the data, regularization
algorithms can take advantage of this structure. For example, if the features
lend themselves to a natural division into different groups, the group lasso
(GL) can be used to select groups of features, rather than individual features
\cite{Yuan2006-ky}. The Sparse Group Lasso (SGL) elaborates on this idea by
providing control both of group sparsity, as well as overall sparsity of the
solutions \cite{simon2013sgl}. Because the features measured with tractomery
lend themselves to grouping based on the tracts from which each measurement is
taken, GL and SGL provide a useful tool linear model fitting in problems of this
form. Using linear regression as the basis for analysis is useful for both
regression tasks, in which a continuous measure (e.g., age) is modeled, as well
as for classification tasks, where a discrete property is modeled (e.g., whether
an individual has a particular disease). Moreover, use of SGL helps derive
informative properties of the human white matter, both in cases where the
underlying causes of variance are sparsely concentrated (e.g., in one particular
white matter tract) and in cases where biological substrates of differences
between individuals are more widespread. This is because SGL is controlled by
meta-parameters that set the degree to which group sparsity and overall sparsity
are applied. These metaparameters can be set empirically, based on the data,
through a process of nested cross-validation. We demonstrate these properties of
the algorithm in data from previously published studies
\cite{sarica2017corticospinal, yeatman2014lifespan}.

### 2019-12-12 reports/manuscript/sections/results.tex

@@

Using data from a previous study of the corticospinal tract (CST) profile and
ALS\cite{sarica2017corticospinal}, we tested the performance of SGL in a
classification setting. The previous study predicted ALS status with a mean
accuracy of 80\% using a random forest algorithm based on a priori selection of
features within the corticospinal tract. SGL delivers competitive predictive
performance (mean 93\% $\pm$ 2\% accuracy, 0.978 $\pm$ 0.006 ROC AUC) without
the need for a priori feature engineering. The results of the classification
prediction are shown in Fig~\ref{fig:class-results} with ``ground-truth'' ALS
status separated into columns, and predicted ALS status encoded by color. In
addition to this classification performance, SGL also identifies the white
matter tracts most important for ALS classification. The relative importance of
white matter features is captured in the $\beta$ coefficients from
Eq~\eqref{eq:sgl}. Fig~\ref{fig:class-profiles} depicts these coefficients along
the right CST, plotted over the FA values for the control and ALS subject
groups. We find that SGL selects
FA metrics in the corticospinal tract and particualrly in the right
corticospinal tract as most important to ALS classification, confirming previous
findings\cite{van2011upper, toosy2003diffusion, sarica2014tractography,
sage2007quantitative, sage2009quantitative, karlsborg2004corticospinal,
ellis1999diffusion, cosottini2005diffusion, ciccarelli2009investigation,
abe2010voxel} and identifying exactly the portions of the brain that were
selected \emph{a priori} in the previous study from which we collected our
data\cite{sarica2017corticospinal}.

We also find that mislabelled subjects correspond to outlying features in the
CST. Fig~\ref{fig:class-errors} depicts the group FA values along with
FA values of mislabelled subjects, two false negatives and one false-positive.
The false negative classifications have high FA in one of the two sections of
the CST where $||\hat{\beta}|| > 0$ in Fig~\ref{fig:class-profiles}.
The false positive subject has an FA that oscillates between the two group
means. Thus, the SGL method fails comprehensibly.

\begin{figure}[h]
    \centering
        \includegraphics[width=0.95\textwidth]{classification_subjects_profiles.pdf}
    \caption{{\bf Model mis-classifications correspond to features identified by
       the model}. The FA in CST of individuals that are mis-classified by the
       model is compared the group FA (with shaded are indicating standard error
       of the mean). False negative classifications (individuals that have ALS,
       but are classified as patients) correspond to high FA either in one of
       the two regions of large $||\hat{\beta}||$ in
       Fig~\ref{fig:class-profiles}. The false positive classification has
       an FA profile that oscillates between the two group means.
   }
    \label{fig:class-errors}
\end{figure}

>>>>

We developed a method for analyzing dMRI tractometry data with SGL. We
demonstrate the use of this method on two previously published datasets in both
a classification setting and a regression setting.

Using data from a previous study of the corticospinal tract (CST) profile and
ALS\cite{sarica2017corticospinal}, we tested the performance of SGL in a
classification setting. The previous study predicted ALS status with a mean
accuracy of 80\% using a random forest algorithm based on a priori selection of
features within the corticospinal tract. SGL delivers competitive predictive
performance (mean 93\% $\pm$ 2\% accuracy, 0.978 $\pm$ 0.006 ROC AUC) without
the need for a priori feature engineering. The results of the classification
prediction are shown in Figure \ref{fig:class-results} with ``ground-truth'' ALS
status separated into columns, and predicted ALS status encoded by color. In
addition to this classification performance, SGL also identifies the white
matter tracts most important for ALS classification. The relative importance of
white matter features is captured in the $\beta$ coefficients from
Eq~\eqref{eq:sgl}. Figure \ref{fig:class-profiles} depicts these coefficients along
the right CST, plotted over the FA values for the control and ALS subject
groups. We find that SGL selects FA metrics in the corticospinal tract and
particularly in the right corticospinal tract as most important to ALS
classification, confirming previous findings\cite{van2011upper,
toosy2003diffusion, sarica2014tractography, sage2007quantitative,
sage2009quantitative, karlsborg2004corticospinal, ellis1999diffusion,
cosottini2005diffusion, ciccarelli2009investigation, abe2010voxel} and
identifying the portions of the brain that were selected \emph{a priori} in the
previous study from which we collected our data\cite{sarica2017corticospinal}.

Analyzing the ways in which the model mislabels individuals may also provide
insight. We found that mislabelled subjects are outliers relative to their group
with respect to diffusion features of the CST. Figure \ref{fig:class-errors}
depicts the group FA values along with FA values of mislabelled subjects, two
false negatives and one false-positive. The false negative classifications have
high FA in one of the two sections of the CST where $||\hat{\beta}|| > 0$ in
Figure \ref{fig:class-profiles}. The false positive subject has an FA that
oscillates between the two group means. Thus, the SGL method fails
comprehensibly.

\begin{figure}[h]
    \centering
        \includegraphics[width=0.95\textwidth]{classification_subjects_profiles.pdf}
    \caption{{\bf Model mis-classifications correspond to features identified by
       the model}. The FA in CST of individuals that are mis-classified by the
       model is compared the group FA (with shaded are indicating standard error
       of the mean). False negative classifications (individuals that have ALS,
       but are classified as patients) correspond to high FA either in one of
       the two regions of large $||\hat{\beta}||$ in
       Figure \ref{fig:class-profiles}. The false positive classification has
       an FA profile that oscillates between the two group means.
   }
    \label{fig:class-errors}
\end{figure}

@@
